Parsing Postfix log files

### 2007-11-07 doc/logparser-acronyms.tex
More corrections and improvements.

>>>>

\newacronym{FQDN}{Fully Qualified Domain Name}{description={
    An FQDN is a hostname plus domain name, e.g.
        example.com     is a domain name    \newline
        www             is a hostname       \newline
        www.example.com is a FQDN
}}

\newacronym{DFA}{Deterministic Finite Automata}{description={
    XXX EXPLAIN DFA\@.  SHOULD IT BE NFA\@?
}}

\newacronym{DNS}{Domain Name System}{description={
    The DNS converts between hostnames (www.example.com) and IP addresses
    (10.1.2.3).
}}

### 2007-11-07 doc/logparser.tex
More corrections and improvements.

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects mail when the triple of (sender, recipient, remote IP
        address) is unknown; on second and subsequent delivery attempts
        from that triple the mail will be accepted.  The assumption is that
        maintaining a list of failed addresses and retrying after a
        temporary failure is uneconomical for a spammer, but that a
        legitimate mail server must retry.  Sadly spammers are using
        increasingly complex and well written programs to distribute spam,
        frequently using an \ISP{} provided \SMTP{} server from a
        compromised machine on the \ISP{}'s network.  Greylisting is slowly
        becoming less useful, but it does block a large percentage of spam
        mail at the moment.

@@

The database is an integral part of the parser: it stores the rules and the
results of applying those rules to Postfix log files.  Understanding the
database schema is important in understanding the actions of the parser,
and essential to developing further applications which utilise the data
gathered.

@@

The database schema can be conceptually divided in two: the rules which are
used to parse log files, and the data saved from the parsing of log files.
Rules have the fields required to parse the log lines, extract data to be
saved, and the action to be executed; they also have several fields which
aid the user in understanding what each rule parses.  The rules are
described in detail in section~\refwithpage{rules} but the fields are
covered in section~\refwithpage{rule attributes}.

@@

    \item [regex] The regular expression\cite{wikipedia-regex, perlre} to
        match the log line against.  The regex will first have several
        keywords expanded: this simplifies reading and writing rules;
        avoids needless repetition of complex regex components; allows the
        components to be corrected and/or improved in one location; and
        makes each regex largely self-documenting.

@@

        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
        \newline \hspace*{2em} server\_hostname and helo.\newline}
        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
        \newline \hspace*{2em} pid\_regex, data and child.}

    \item [result\_data, connection\_data] Sometimes rules need to supply a
        piece of data which isn't present in the log line: e.g.\ setting
        \texttt{smtp\_code} when mail is accepted.  The format and allowed
        variables are the same as for \texttt{result\_cols} and
        \texttt{connection\_cols}, except that arbitrary
        data\footnote{Commas and semi-colons cannot be escaped and thus
        cannot be used.  This is intended for use with small amounts of
        data rather than large, so dealing with escape sequences seemed
        unnecessary.} is permitted on the right hand side of the
        assignment.

@@

    \item [server\_hostname] The hostname of the server, it will be
        \textit{unknown\/} if the IP address could not be resolved to a
        hostname via DNS\@.

@@

    \item [client\_hostname] The hostname of the client, it will be
        \textit{unknown\/} if the IP address could not be resolved to a
        hostname via DNS\@.

@@

        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
        place to log the HELO hostname, there won't be a queueid associated
        with the mail for the first recipient, so that log line cannot be
        associated with the correct mail.  There is guaranteed to be
        queueid when the DATA command has been reached, and thus it will be
        logged by any restrictions taking effect in
        \texttt{smtpd\_data\_restrictions}.

    \item [queueid] The queueid of the mail if the conection represents an
        accepted mail, or \textit{NOQUEUE\/} otherwise.

@@

    \item [connection\_id] A reference to the entry in the connections
        table this results table entry belongs to.

@@

    \item [message\_id] The message-id of the accepted mail, or NULL if no
        mail was accepted.

@@

The table containing the rules used by the parser, and the two tables
containing the data extracted from the Postfix logs were described, with
the purpose of each field discussed in detail.  A clear, comprehensible
schema is essential in using the extracted data; it's more important for
using the data than storing the data, because storing the data is a
write-once operation, whereas utilising the data requires frequent
searching, sorting and manipulation of the data to produce customised
reports and/or statistics.

\section{Adaptable parsing: user defined rules}

@@

The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  The complexity and variation in Postfix's
logs requires similar flexibility in the parser; decoupling the parsing
rules from the associated actions allows new rules to be written and tested
without requiring modifications to the algorithm source code (significantly
lowering the barrier to entry for new or casual users), and greatly
simplifies both algorithm and rules.  It also creates a clear separation of
functionality: rules handle low level details of identifying log lines and
extracting data from a line, whereas the algorithm handles the higher level
details of following the path a mail takes through Postfix, assembling the
required data, etc.

@@

    \item Rules are annotated with the name of a Postfix program, and will
        only be used when parsing log lines produced by that
        program.\footnote{There are also generic rules which are used when
        parsing log lines produced by any Postfix program, but only if
        there are also rules specific to that program, and those rules must
        already have been tried and failed.}  Any given rule will only be
        used to parse a subset of the log lines, and any given log line
        will only be parsed by a subset of the rules.

@@

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's regex, generate an NFA, and compare all NFAs
to determine if any overlap.  The author has not attempted to do this: such
a project would probably qualify for a PhD, and may involve solving the
Halting Problem and circumventing the Church-Turing Thesis.} that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line determines which rule
matches.  Overlapping rules are frequently a requirement, allowing a more
specific rule to match some lines and a more general rule to match the
majority, e.g.\ separating \SMTP{} delivery to a filtering proxy from
\SMTP{} delivery to the rest of the world.  The algorithm provides a
facility for ordering overlapping rules: the priority field in each rule
(defaults to zero).  Rules are tried in order of priority, highest first,
overriding the hits field maintained by the parser.  Negative priorities
may also be useful for catchall rules.

@@

    \item Compare the results of parsing using sorted, shuffled and
        reversed rules.\footnote{See section~\refwithpage{rule efficiency}
        for more details of sorting the rules.}  Parse a number of log
        files using normal sorting, dump a textual representation of the
        connections and results tables, and delete everything from those
        tables.  Repeat with shuffled and reversed sorting.  If there are
        no overlapping rules the tables from each run will be identical;
        differences indicate overlapping rules, and the rule\_ids which
        differ determine which rules overlap.  Unfortunately this method
        cannot prove the absence of overlapping rules; it can detect
        overlapping rules --- but only if there are log lines in the input
        files which match more than one rule.

@@

This example rule matches the message logged by Postfix when it rejects
mail from a sender address where the domain has neither an MX record nor an
A record, i.e.\ mail could not be delivered to the sender's address --- for
full details see~\cite{reject-unknown-sender-domain}.

@@

    \item [program and regex:] If the program in the rule equals the
        program which logged the line the regex will be matched against the
        line; if the match is successful the action will be executed, if not
        the next rule will be tried.  If the program-specific rules don't
        match the log line, the generic rules will be tried in the same
        way.

@@

    \item [queueid:] is used the extract the queueid matched by the regex,
        so that the correct mail can be found by queueid and actions
        performed on it.

    \item [hits, hits\_total and priority:] hits and priority are used in
        ordering the rules (see section~\refwithpage{rule ordering for
        efficiency}); hits\_total is updated at the end of each run but is
        otherwise unused.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
deal with 75 MB log files containing 300,000 log lines (for a small mail
server --- large scale mail servers would have much larger log files on a
daily basis).  When generating the data for the graphs included in
section~\refwithpage{graphs}, 93 log files (totaling 10.08 GB, 60.72
million lines) were each parsed 10 times, the first run discarded and the
remaining 9 runs averaged.  The first run is discarded for two reasons:

@@

Rule ordering was mentioned in section~\refwithpage{rule attributes} and
will be covered in greater detail in this section.  At the time of writing
there are 139 different rules, with the top 10\% matching the vast majority
of the log lines, and the remaining log lines split across the other 90\%
of the rules (as shown in graph~\refwithpage{rule hits}).  Assuming that
the distribution of log lines is reasonably steady over time, program
efficiency should benefit from trying more frequently matching rules before
those which match less frequently.  To test this hypothesis three full test
runs were performed with different rule orderings:

@@

Graphs~\refwithpage{percentage increase of shuffled over normal}
and~\refwithpage{percentage increase of reversed over normal} show the
percentage increase of execution times (mean and standard deviation are
shown after the graph).  Overall this provides a modest but worthwhile
performance increase.

@@

Caching the compiled regexs is quite simple, and is the single most
effective optimisation tried in the parser.

@@

    \item An example log line and the rule matching it was given, followed
        by a description of how the fields in the rule are used both in the
        matching and the subsequent action that is executed.

@@

    \item The final part of this section discusses rule efficiency,
        covering the effects of caching compiled regexs and optimal
        ordering ordering of rules, with reference to the graphs in
        section~\refwithpage{graphs}.

@@

    \item A flow chart showing the paths a mail can take through the
        algorithm.

    \item A description of the full algorithm.

    \item The actions which can be taken by the parser are described in
        detail.

    \item Additional complications which have arisen during the development
        of this parser are described.

@@

>>>>

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects mail when the triple of (sender, recipient, remote IP
        address) is unknown; on second and subsequent delivery attempts
        from that triple the mail will be accepted.  The assumption is that
        maintaining a list of failed addresses and retrying after a
        temporary failure is uneconomical for a spammer, but that a
        legitimate mail server must retry.  Sadly spammers are using
        increasingly complex and well written programs to distribute spam,
        frequently using an \ISP{} provided \SMTP{} server from a
        compromised machine on the \ISP{}'s network.  Greylisting will
        slowly become less useful, but it does block a large percentage of
        spam mail at the moment; the most effective restrictions over the
        93 log files used in testing the parser are shown in
        table~\refwithpage{Summary of rejections}.  Greylisting is
        obviously worth using, at least at the moment, particularly when
        you factor in Greylisting's position as the final restriction which
        a mail must overcome: Greylisting only takes effect for mails which
        have passed every other restriction.

        \begin{table}[ht]
            \caption{Summary of rejections}\label{Summary of rejections}
            \begin{tabular}[]{lr}
                Restriction type    & Total rejections  \\
                Total rejections    & 5283739           \\
                Unknown recipient   & 2338000           \\
                Unknown sender      & 864148            \\
                \RBL{}s             & 807185            \\
                Greylisting         & 497244            \\
                Non-\FQDN{} HELO    & 403652            \\
            \end{tabular}
        \end{table}

@@

The database is an integral part of the parser: it stores the rules and the
data gleaned by applying those rules to Postfix log files.  Understanding
the database schema is important in understanding the actions of the
parser, and essential to developing further applications which utilise the
data gathered.

@@

The database schema can be conceptually divided in two: the rules which are
used to parse log files, and the data saved from the parsing of log files.
Rules have the fields required to parse the log lines, extract data to be
saved, and the action to be executed; they also have several fields which
aid the user in understanding what each rule parses.  The rules are
described in detail in section~\refwithpage{rules} but the fields are
covered here.

@@

    \item [regex] The regular expression~\cite{wikipedia-regex, perlre} to
        match the log line against.  The regex will first have several
        keywords expanded: this simplifies reading and writing rules;
        avoids needless repetition of complex regex components; allows the
        components to be corrected and/or improved in one location; and
        makes each regex largely self-documenting.

@@

        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
        \newline \hspace*{2em} server\_hostname} and \texttt{helo.\newline}
        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
        \newline \hspace*{2em} pid\_regex, data} and \texttt{child.}

    \item [result\_data, connection\_data] Sometimes rules need to supply a
        piece of data which isn't present in the log line: e.g.\ setting
        \texttt{smtp\_code} when mail is accepted.  The format and allowed
        variables are the same as for \texttt{result\_cols} and
        \texttt{connection\_cols}, except that arbitrary
        data\footnote{Commas and semi-colons cannot be escaped and thus
        cannot be used.  This is intended for use with small amounts of
        data rather than large amounts, so dealing with escape sequences
        seemed unnecessary.} is permitted on the right hand side of the
        assignment.

@@

    \item [server\_hostname] The hostname of the server, it will be
        \texttt{unknown} if the IP address could not be resolved to a
        hostname via DNS\@.

@@

    \item [client\_hostname] The hostname of the client, it will be
        \texttt{unknown} if the IP address could not be resolved to a
        hostname via DNS\@.

@@

        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
        place to log the HELO hostname, there won't be a queueid associated
        with the mail for the first recipient, so that log line cannot be
        associated with the correct mail.  There is guaranteed to be
        queueid when the DATA command has been reached, and thus it will be
        logged by any restrictions taking effect in
        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
        specifying a HELO-based restriction in
        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
        correctly.

    \item [queueid] The queueid of the mail if the connection represents an
        accepted mail, or \texttt{NOQUEUE} otherwise.

@@

    \item [connection\_id] A reference to the row in the connections table
        this result's table row belongs to.

@@

    \item [message\_id] The message-id of the accepted mail, or
        \texttt{NULL} if no mail was accepted.

@@

The table containing the rules used by the parser, and both tables
containing the data extracted from the Postfix logs were described, with
the purpose of each field discussed in detail.  A clear, comprehensible
schema is essential when using the extracted data; it's more important when
using the data than when storing it, because storing the data is a
write-once operation, whereas utilising the data requires frequent
searching, sorting and manipulation of the data to produce customised
reports and/or statistics.

\section{Parsing rules}

@@

The complexity and variation in Postfix's logs requires similar flexibility
in the parser, which is split into two parts: the parsing algorithm and the
parsing rules which are applied to the lines.  Decoupling the parsing rules
from the associated actions allows new rules to be written and tested
without requiring modifications to the algorithm source code (significantly
lowering the barrier to entry for new or casual users who need to parse new
log lines), and greatly simplifies both algorithm and rules.  Decoupling
also creates a clear separation of functionality: rules handle low level
details of identifying log lines and extracting data from a line, whereas
the algorithm handles the higher level details of following the path a mail
takes through Postfix, assembling the required data before storing it,
dealing with complications arising, etc.

@@

    \item Rules are annotated with the name of a Postfix program, and will
        only be used when parsing log lines produced by that
        program.\footnote{There are also generic rules which are used when
        parsing log lines produced by any Postfix program, but only if
        there are also rules specific to that program, and those rules must
        already have been tried and failed on the current line.}  Any given
        rule will only be used to parse a subset of the log lines, and any
        given log line will only be parsed by a subset of the rules.

@@

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's regex, generate a \DFA{} from it, and examine
every \DFA{} to determine if any overlap.  The author has not attempted to
do this: such a project by itself would probably qualify for a PhD, and may
involve solving the Halting Problem and circumventing the Church-Turing
Thesis.  XXX REFERENCES FOR HALTING PROBLEM AND CHURCH-TURING\@.} that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line is unspecified, and
the first matching rule wins.  Overlapping rules are frequently a
requirement, allowing a more specific rule to match some lines and a more
general rule to match the majority, e.g.\ separating \SMTP{} delivery to a
specific sites from \SMTP{} delivery to the rest of the world.  The
algorithm provides a facility for ordering overlapping rules: the priority
field in each rule (defaults to zero).  Rules are sorted by priority,
highest first, and then rules with the same priority are sorted by the
number of successful matches when parsing the previous log file.  Negative
priorities may be useful for catchall rules.

@@

    \item Compare the results of parsing using sorted, shuffled and
        reversed rules.\footnote{See section~\refwithpage{rule efficiency}
        for more details of sorting the rules.}  Parse a number of log
        files using normal sorting, dump a textual representation of the
        connections and results tables, and delete everything from those
        tables.  Repeat with shuffled and reversed sorting.  If there are
        no overlapping rules the tables from each run will be identical;
        differences indicate overlapping rules.  Which rules overlap can be
        determined by examining the differences in the tables: each result
        contains the id of the rule which created it, so the rule
        referenced in the normal table overlaps with the rule referenced in
        the reversed table.  Unfortunately this method cannot prove the
        absence of overlapping rules; it can detect overlapping rules, but
        only if there are log lines in the input files which match more
        than one rule.

@@

This example rule matches the message logged by Postfix when it rejects
mail from a sender address because the appropriate \DNS{} entries are
missing, i.e.\ mail could not be delivered to the sender's address (for
full details see~\cite{reject-unknown-sender-domain}).

@@

    \item [program and regex:] If the program in the rule equals the
        program which logged the line the regex will be tried against the
        line; if the match is successful the action will be executed, if
        not the next rule will be tried.  If the program-specific rules
        don't match the log line, the generic rules will be tried in the
        same way.

@@

    \item [queueid:] The index of the field in the regex which captured the
        queueid, or zero if the line does not contain a queueid.  This
        allows the correct mail can be found by queueid and actions
        performed on it.

    \item [hits, hits\_total and priority:] hits and priority are used in
        ordering the rules (see section~\refwithpage{rule ordering for
        efficiency}); hits is set to the number of successful matches at
        the end of the parsing run, and hits\_total has hits added to it,
        but is otherwise unused by the algorithm.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
deal with 75 MB log files containing 300,000 log lines (for a small mail
server --- large scale mail servers would have much larger log files on a
daily basis).  When generating the data for the graphs included in
appendix~\refwithpage{graphs}, 93 log files (totaling 10.08 GB, 60.72
million lines) were each parsed 10 times, the first run discarded, and the
remaining 9 runs averaged.  The first run is discarded for two reasons:

@@

Rule ordering was mentioned in section~\refwithpage{rule attributes} and
will be covered in greater detail in this section.  At the time of writing
there are 139 different rules, with the top 10\% matching the vast majority
of the log lines, and the remaining log lines split across the other 90\%
of the rules (as shown in graph~\refwithpage{rule hits graph}).  Assuming
that the distribution of log lines is reasonably steady over time, program
efficiency should benefit from trying more frequently matching rules before
those which match less frequently.  To test this hypothesis three full test
runs were performed with different rule orderings:

@@

Graphs~\refwithpage{percentage increase of shuffled over normal}
and~\refwithpage{percentage increase of reversed over normal} show the
percentage increase of execution times (mean and standard deviation are
shown in table~\refwithpage{Regex caching/discarding with different groups
of log files}).  Overall this provides a modest but worthwhile
performance increase.

@@

Caching the compiled regexs is quite simple, and is the single most
effective optimisation implemented in the parser.

@@

    \item An example log line and the rule matching it illustrated a
        description of how the fields in the rule are used both in the
        matching phase and the subsequent action that is executed.

@@

    \item The final part of this section discusses rule efficiency,
        covering the effects of caching compiled regexs and optimal
        ordering of rules, with reference to the graphs in
        appendix~\refwithpage{graphs}.

@@

    \item A flow chart showing common paths a mail can take through the
        algorithm.

    \item An explanation of the paths shown in the flow chart.

    \item The actions the parser makes available to rules are covered in
        detail.

    \item Additional complications which have arisen during the development
        of this parser are documented, including their solutions and where
        those solutions are implemented in the algorithm

@@

\glossary{name={Epoch},description={
    Most operating systems store the current time and timestamps of files
    etc.\ as seconds elapsed since the epoch, the beginning of time as far
    as the operating system is concerned.  On Unix and Unix-derived systems
    the epoch is 1970/01/01 00:00:00; on other operating systems it may be
    different.
}}

\glossary{name={NULL},description={
    NULL is a special term used in \SQL{} databases indicating the absence
    of data for the field.
}}
